---
title: "Notes while learning Bayesian inference"
author: "Dave Bridges"
date: "2025-12-13"
editor: source
format: 
  html:
    toc: true
    toc-location: right
    keep-md: true
    code-fold: true
    code-summary: "Show the code"
    fig-path: "figures/"
theme: journal
bibliography: references.bib
execute:
  echo: true
  warning: false
---

```{r global_options}
# hide this code chunk
#| echo: false
#| message: false

# defines the se function
se <- function(x) {
  sd(x, na.rm = TRUE) / sqrt(length(x))
}

#load these packages, nearly always needed
library(tidyverse)
library(knitr)

# sets maize and blue color scheme
color_scheme <- c("#00274c", "#ffcb05")
```

These are notes as I work through and understand the notes from BIOSTAT 682

Bayes theoreom says

$$Pr(\theta|y) = \frac{Pr(y|\theta)\cdot Pr(\theta)}{Pr(y)}$$

Where:

- $Pr(\theta|y)$ is the posterior probability, given the data $y$
- $Pr(y|\theta)$ is the likelihood function (the probability of getting data $y$ presuming the current hypothesis $\theta$ is correct)
- $Pr(\theta)$ is the prior probability
- $Pr(y)$ is probability of observing the data

## Conjugate Priors

Conjugate priors for a distribution are chosen so that the prior and the posterior have the same form.  According to Bayes' theorem $Pr(\theta | y) \propto Pr(y|\theta) \cdot \Pr(\theta)$ where both $Pr(\theta)$ (the prior) and $\Pr(\theta|y)$ (the posterior) are distributions over the parameter $\theta$.  This allows us to do Bayesian updating of models as new data arrives without the computationally expensive computation of the product of the likelihood and the prior.

## Single Parameter Models

The motivating example here is that there was an incidence of cancer wherein 8 cancers appeared out of a total of 145 people.  The question is whether this incidence is greater than the expected incidence of 4.458% from [@CancerStatisticsNCI2015]

Using the binomial distribution probability mass function for a binomial distribution

$$P(Y = y | n, \theta) = \binom{n}{y} \theta^y (1 - \theta)^{n-y}$$
In this nomenclature $\binom{n}{y}$ expands out to $\frac{n!}{y! \times (1-y)!}$ where $y$ is the number of successes, $n$ is the number of trials.  In the overall cuntion $\theta$ is the probability of success on each trial.  


This can also be denoted as $Y \sim \text{Binomial}(n, \theta)$ where $n$ is the number of trials and $\theta$ is the probability of success on each trial, Using this we can calculate the likelihood of observing 8 or more cancers out of 145 people given the expected incidence rate of 4.44% and some other probabilities

The hypothesis being tested are that:

- $H_A: \theta = 0.03$
- $H_B: \theta = 0.04$ 
- $H_C: \theta = 0.05$
- $H_D: \theta = 0.06$


In terms of prior probabilities, we can assign prior probabilities to each of the hypotheses above.  If we think that $H1$ is 50% likely, and $H_{A-D}$ are equally likely after that.

Likelihoods are calculated from the probability mass function for the binomal function $Y \sim \text{Binomial}(n, \theta)$ 

 can be calculated for each tested hypothesis ($H_x$ in comparason to all possible hypotheses $Hi$) via $P(H_x \mid y_o = 8) = \frac{P(H_x) \times, P(y_o = 8 \mid H_x)}{\sum_{i=A}^{D} P(H_i) \times, P(y_o = 8 \mid H_i)}$.  In simplified terms this is the prior probability of the hypothesis, times its likelihood divided by the sum of all prior probabilities times their likelihoods.

```{r lecture-2-cancer}
n <- 145
cases <- 8
theta <- c(0.03, 0.04, 0.05, 0.06)
priors <- c(0.5, rep(0.5/3,3))

data.frame(hypothesis = theta,
           priors=priors) |>
  mutate(likelihood = dbinom(cases, size = n, prob = hypothesis)) |> #this does not give the same results as the notes probably an approximation
  mutate(rel.likelihood = likelihood / min(likelihood)) |>
  mutate(posterior = priors * likelihood / sum(priors * likelihood)) |>
  kable(caption=paste0("Probability of observing ", cases, " cases out of ",n, " trials"),
        digits=c(3,3,2))

#this does not give the same results as the notes
```

This means that hypotheses 4 and 5 explain the data approximately the same times better than hypothesis 1.

#### The likilihood principal

There is an articulation of the *Likelihood Principal*: once $Y$ has been observed we now note it as $Y = y_o$ , and no other value of Y matters. We should treat as $L(\theta|y_o) = Pr(Y = y_o | \theta)$ as the only likelihood function of $\theta$ and no other aspects of the experiment or data matter for inference.

#### The conjugate prior for a binomial distribution is a beta distribution

What if the prior probabilities are continuous, if we model them as $\theta = Beta(\alpha,\beta)$.  The density of this prior is given by:

$$ \pi(\theta \mid \alpha, \beta) = \frac{\Gamma(\alpha + \beta)}{\Gamma(\alpha)\Gamma(\beta)} \theta^{\alpha - 1} (1 - \theta)^{\beta - 1} $$

In this case $\frac{\Gamma(\alpha + \beta)}{\Gamma(\alpha)\Gamma(\beta)}$ is a normalizing factor that ensures probabilities intgrate to one and exist between zero and one.

The posterior density is therefore proprotional to to the prior times the likelihood $\pi(\theta \mid y) \propto \pi(\theta) \times p(y \mid \theta)$.  This expands out to this $\pi(\theta \mid y) \propto \theta^{\alpha - 1} (1 - \theta)^{\beta - 1} \times \theta^y (1 - \theta)^{n - y}$ droppping the constants for now.  By combining the exponents we get $\pi(\theta \mid y) \propto \theta^{(\alpha + y) - 1} (1 - \theta)^{( \beta + n - y) - 1}$.  This is the kernel (un-normalized density) of a Beta distribution with updated parameters $\alpha' = \alpha + y, \quad \beta' = \beta + n - y$

To get the proper density, we need to insert the Beta normalizing constant with the new parameters $\pi(\theta \mid y) = \frac{\Gamma(\alpha + y + \beta + n - y)}{\Gamma(\alpha + y)\Gamma(\beta + n - y)} \theta^{\alpha + y - 1} (1 - \theta)^{\beta + n - y - 1}$ which simplifies to
$\theta \mid y \sim \operatorname{Beta}(y + \alpha, \, n - y + \beta)$.  This is the posterior probability distribution.

To calculate the mean and variance, for any $\operatorname{Beta}(a, b)$ random variable:$\mathbb{E}[\theta] = \frac{a}{a + b}, \quad \operatorname{Var}(\theta) = \frac{ab}{(a + b)^2 (a + b + 1)}$ we can plug in $a = y + \alpha$, $b = n - y + \beta$: $\mathbb{E}[\theta \mid y] = \frac{y + \alpha}{n + \alpha + \beta}$ (this is a weighted average between the observed proportion $y/n$ and the prior “pseudo-mean” $\alpha/(\alpha + \beta)$, with more weight on the data when n is large) $\operatorname{Var}(\theta \mid y) = \frac{(y + \alpha)(n - y + \beta)}{(n + \alpha + \beta)^2 (n + \alpha + \beta + 1)}$

To summarize if our priors are $Beta(\alpha, \beta)$ and our posterior probabilities are $P(\theta|Y=y_o=8)=Beta(y_o + α, n − y_o + β)$


```{r lecture-2-continuous-prior}
#mean = alpha / (alpha + beta) so to guestimate 0.444 is 4/85
alpha <-  1
beta <- 1

range <- seq(0,1,by=0.001)
prior <- dbeta(range,shape1=alpha,shape2=beta)
posterior <- dbeta(range,shape1=cases + alpha, shape2=n - cases + beta)
data.beta <- data.frame(range=range,
                        prior=prior,
                        posterior=posterior) |>
  pivot_longer(cols=c("prior","posterior"),
               names_to="distribution",
               values_to="value")

ggplot(data.beta,aes(y=value,x=range,col=distribution)) +
  geom_line() +
  labs(title=paste0("Distribution - Beta(",round(alpha,2),",",round(beta,2),")"),
       x="Value",
       y="Likelihood") +
  theme_classic(base_size=16) +
  theme(legend.position = c(0.85,0.85)) 

#these are defined from the conjugate prior/posterior relationship
prior_mean = alpha/(alpha+beta)
prior_ci_lower = qbeta(0.025, alpha, beta)
prior_ci_upper = qbeta(0.975, alpha, beta)
posterior_mean = (alpha + cases)/(alpha+cases+n-cases+beta)
posterior_ci_lower = qbeta(0.025, (alpha + cases), (cases+n-cases+beta))
posterior_ci_upper = qbeta(0.975, (alpha + cases), (cases+n-cases+beta))

data.frame(Distribution=c("Prior","Posterior"),
           Mean=c(prior_mean,posterior_mean),
           CI_Lower=c(prior_ci_lower,posterior_ci_lower),
           CI_Upper=c(prior_ci_upper,posterior_ci_upper)) |>
  kable(digits=3,
        caption="Summary statistics for prior and posterior distributions")
```

By this model the posterior probability that the incidence is greater than 4.44% is `r round(pbeta(0.0444,alpha+cases,n-cases+beta, lower.tail = F)*100,1)`%.

###### Example of Beta as a Conjugate Prior

Imagine there is a mouse that has a rare phenotype, we observe $n=20$ mice, and find $y=4$ have this phenotype. The probability of this occuring can be denoted as $Y | p \sim Bin(20,p)$.  Before noticing this, we believe the phenotype is uncommon but i have low certainty on this hypothesis, given as $p \sim Beta(2,20)$. This was selected because in $Beta(\alpha,\beta)$ this means that

- $\alpha-1$ is how often its observed (successes).  By that $\alpha=\text{successes}+1$
- $\beta-1$ is the number of times not observed (failures).  By that $\beta=\text{failures}+1$
- $\alpha + \beta$ is the strength of this conviction
- The mean estimate is $E[p]=\frac{\alpha}{\alpha + \beta}$

So this prior is formulating on the expectation that out of 18 examples, I would have expected one example of this phenotype.  This corresponds to an estimate of 10% $\frac{2}{2+19}$.  Here are three possible visualizations of priors for this example, presuming one success out of 18, 5, or 30 trials

```{r binomial-conjugate-example-prior-plot}
prior.data.frame <- data.frame(probability=seq(0,1,by=0.01)) |>
  expand_grid(observed=1, attempts=c(20,5,30)) |>
  mutate(alpha = observed + 1,
         beta = attempts - observed + 1) |>
  mutate(prior = paste0("Beta(",alpha,",",beta,")")) |>
  mutate(density=dbeta(probability,alpha,beta)) 

ggplot(prior.data.frame, aes(y=density,x=probability,col=prior)) +
  geom_line() +
  theme_classic(base_size = 16) +
  theme(legend.position=c(0.8,0.8))
```

The posterior is the likelihood times prior is $L(p|y_o=4) \times \pi(p)$.  This becomes $(p∣y_o) \sim Beta(\alpha+Y_o,\beta+n−y_o)$.  For this example this computes to $Beta(2+4,20+20-4)=Beta(6,36)$


```{r beta-example-prior-posterior}
data.frame(probability=seq(0,1,by=0.01)) |>
  mutate(prior = dbeta(probability,2,18),
         posterior = dbeta(probability,6,36)) |>
  pivot_longer(c(prior,posterior),names_to="type",values_to = "density") |>
  ggplot(aes(y=density,x=probability,col=type)) +
  geom_line() +
  geom_vline(xintercept=4/20, lty=2, col="grey") +
  annotate(
    geom  = "text",
    x     = 4/20+0.01,           # ← your chosen x-position
    y     = 8,            # ← your chosen y-position
    label = "Observed",
    hjust = 0,            # 0 = left justified (most important part)
    vjust = 0.5,          # optional: 0.5 = vertically centered
    size  = 5, color = "grey"
  ) +
  theme_classic(base_size=16) +
  theme(legend.position=c(0.8,0.8))
```

Given this result of a posterior probability of $Beta(6,36)$ this corresponds to a mean probability of $\frac{6}{6+36}=\frac{6}{42}=$ `r 6/(6+36)`.  The variance is $Var(p)=\frac{\alpha \times \beta}{(\alpha + beta)^2\times(\alpha + \beta + 1)}= \frac{6\times 36}{(6+36)^2\times(6+36+1)}=$ `r (6*36)/((6+36)^2*(6+36+1))`.  The posterior mode is $MAP=\frac{\alpha-1}{\alpha + \beta - 2}=$ `r (6-1)/(6+36-2)`.

###### Nutrition Example

You are studying a rare adverse event from a new dietary supplement.  From prior literature, events are uncommon, but not impossible.  You now run a small pilot study. You observe 7 adverse events out of 50 participants.

I think a reasonable prior probability for adverse events would be something like 1% with reasonable certainty (perhaps 1000 prior data points with 10 being affected, so $\alpha=\text{successes}+1=11$ 990 being unaffected so $\beta=\text{failures}+1=991$), so i think a prior distribution of $Beta(11,991)$ makes sense.  

```{r second-beta-example}
observed <- 7
tested <- 50
alpha <- 18
beta <- 1037
```

By this logic, the posterior probability should be $Beta(\alpha+y_o,beta+n-y_o)$ or $Beta(18,1037)$.  This computes to a posterior mean of `r alpha/(beta+alpha)` [`r qbeta(c(0.025, 0.975), alpha, beta)`] in comparison to an observed maximum likelihood estimator (MLE) of `r observed/tested`.  The modal observation is $MAP=\frac{\alpha-1}{\alpha + \beta - 2}=$ or `r (alpha-1)/(alpha+beta-2)` with a variance of $\frac{\alpha \times \beta}{(\alpha + beta)^2\times(\alpha + \beta + 1)}$ or `r (alpha*beta)/((alpha+beta)^2*(alpha+beta+1))`.  This is visualized below, illustrating the effect of a strong prior

```{r second-beta-example-prior-posterior}
data.frame(probability=seq(0,0.2,by=0.001)) |>
  mutate(prior = dbeta(probability,11,991),
         posterior = dbeta(probability,18,1037)) |>
  pivot_longer(c(prior,posterior),names_to="type",values_to = "density") |>
  ggplot(aes(y=density,x=probability,col=type)) +
  geom_line() +
  geom_vline(xintercept=observed/tested, lty=2, col="grey") +
  annotate(
    geom  = "text",
    x     = observed/tested+0.01,           # ← your chosen x-position
    y     = 8,            # ← your chosen y-position
    label = "Observed",
    hjust = 0,            # 0 = left justified (most important part)
    vjust = 0.5,          # optional: 0.5 = vertically centered
    size  = 5, color = "grey"
  ) +
  theme_classic(base_size=16) +
  theme(legend.position=c(0.8,0.8))
```

##### Proper and Improper Distributions

A **proper probability distribution** integrates to 1, this means if you add up the sum of all possible outcomes, you get unity.  Bayes theorom assumes proper posterior probability distributions.  This means that for discrete distributions like the bimomal $\sum \text{probabilities}=1$ and for continuous distributions $\int \text{density}=1$ (discreted distributions give exact probabilities for a number whereas continuous distributions give density between specific ranges).

#### Continuous Distributions

The **central limit theorem** justifies using a normal distribution for many casses.  This means that for a binomial distribution for example, that as n becomes large, the binomial distribution becomes a normal distribution.  The CLT says that averages and estimators can become normal even if the underlying data (*e.g.* binomal, bernoulli, poisson) is not.

For a normal distribution $N(\mu, \sigma^2)$ where $\mu$ is the mean and $\sigma^2$ is the variance (or the $SE^2$).  

##### Specifying a Prior for the Standard Error From the Conjugate Prior of the Normal Distribution

For a Normal likelihood with known mean $\mu$, the conjugate prior for the variance is the inverse-gamma distribution: $\sigma^2 \sim InvGamma(\alpha,\beta)$ where $\alpha$ is the strength of the prior, approximately $\alpha \approx \frac{df_{prior}}{2}$ and $\beta$ is the prior scale, or approximately $\beta \approx \frac{df_{prior} \times s^2_0}{2}$ where $s^2_0$ is the prior guess of the variance.  Key equations:

- For the variance $E(\sigma^2)=\frac{\beta}{\alpha-1}$ and $Var(\sigma^2)=\frac{\beta^2}{(\alpha-1)^2(\alpha-2)}$ 
- The posterior mode is $\frac{\beta}{\alpha+1}$.
- With posterior updating (given a new degree of freedom $\frac{n}{2}$ and sum of squares $\frac{S}{2}$), $\sigma^2|y \sim InvGamma(\alpha + \frac{n}{2},\beta + \frac{S}{2})$

If we use precision $\frac{1}{\sigma^2}$ then the conjugate prior for variance is the Gamma distribution with the same meaning for $\alpha$ and $\beta$ except that $\beta$ is now a rate rather than a scale.

###### Example for Estimating Prior and Posterior Variance

You are studying a biomarker that is known (from calibration experiments) to have mean $\mu=100$.  You are uncertain about the variance $\sigma^2$.  You have this prior information:

- From historical data, you believe that the standard deviation is around 5 units (so $\sigma^2=25$)
- This belief is roughly equivalent to 20 prior observations

Based on this $\alpha$ should be approximately $\frac{20}{2}=10$ and $\beta$ should be approximately $\beta \approx \frac{df_{prior} \times s^2_0}{2} = \frac{20 \times 5^2}{2}=250$.  Therefore the prior for the variance should be described as $\sigma^2 \sim InvGamma(10,250)$.  This gives a prior mean of $E(\sigma^2)=\frac{\beta}{\alpha-1}=\frac{250}{10-1}=$ `r 250/(10-1)` with a variance of `r 250^2/((10-1)^2*(10-2))`.

```{r new-measurements}
new.values <- c(101,96,98,103,99,100,97,102,104,95)
```
You now collect 10 new measurements `r paste(new.values,sep=",")`.  The sum of squares of these new values is `r sum((new.values-100)^2)`.  The posterior distribution of $\sigma^2$ is therefore 

$$\sigma^2|y \sim InvGamma(\alpha + \frac{n}{2},\beta + \frac{S}{2})$$

$$ \sigma^2|y \sim InvGamma(10+\frac{10}{2},250+\frac{85}{2})=InvGamma(15,292.5)$$ 

This is visualized as

```{r continuous-prior-example}
alpha_prior <- 10
beta_prior <- 250

alpha_posterior <- alpha_prior + length(new.values)/2
beta_posterior <- beta_prior + sum((new.values-100)^2)/2

library(invgamma)
data.frame(variance=seq(0,100,by=1)) |>
  mutate(prior=dinvgamma(variance,alpha_prior,beta_prior),
         posterior = dinvgamma(variance,alpha_posterior,beta_posterior)) |> 
  pivot_longer(c(prior,posterior),
               names_to = "Distribution",
               values_to = "density") |>
  ggplot(aes(y=density,x=variance,col=Distribution)) +
  geom_line() +
    geom_vline(xintercept=25, lty=2, col="grey") +
  annotate(
    geom  = "text",
    x     = 25+1,           # ← your chosen x-position
    y     = 0.07,            # ← your chosen y-position
    label = "Initial Estimate",
    hjust = 0,            # 0 = left justified (most important part)
    vjust = 0.5,          # optional: 0.5 = vertically centered
    size  = 5, color = "grey"
  ) +
  theme_classic(base_size = 16) +
  theme(legend.position=c(0.8,0.8))
```
The updated $\sigma^2$ values are then $E(\sigma^2)=\frac{\beta}{\alpha-1}$,  $Var(\sigma^2)=\frac{\beta^2}{(\alpha-1)^2(\alpha-2)}$, and $\text{Mode}=\frac{\beta}{\alpha+1}$ so `r beta_posterior/(alpha_posterior-1)` [`r qinvgamma(c(0.025,0.975), alpha_posterior, beta_posterior)`] with a modal value of `r beta_posterior/(alpha_posterior+1)`. This is compared to the prior estimates which were `r beta_prior/(alpha_prior-1)` [`r qinvgamma(c(0.025,0.975), alpha_prior, beta_prior)`] and a modal value of `r beta_prior/(alpha_prior+1)`.  As you can see from the figure the posterior shifted left with the updated data that had smaller variance than expected.  The updated standard deviation is `r sqrt(beta_posterior/(alpha_posterior-1))`[`r sqrt(qinvgamma(c(0.025,0.975), alpha_posterior, beta_posterior))`] compared to 5.  The probability that the value is below 5 is `r pgamma(1/25, shape = alpha_posterior, rate = beta_posterior, lower.tail=FALSE)`.  Note for this case we have to use Gamma rather than Inverse-Gamma due to point arithmetic issues with large arguments in `pinvgamma`.

##### Poisson Distributions

Poisson distributions work well for independent counts data (*e.g.* 0, 1, 2, 3 ...) and are described by $\text{Poisson}(\gamma)$, which has a probability mass function of $Pr(Y=y_o|\lambda)=\frac{\lambda^{y_o} \cdot e^{-\lambda}}{y_o!}$ where $\lambda$ describes the expected rate of a certain number events and $y_o$ is observed the number of events [@poisson1837probabilite].  $\lambda$ is both the mean and variance of this distribution, with the modal value given by the largest integer less than or equal to $\lambda$ or $\text{floor}(\lambda)$.

The conjugate prior for a Poisson distribution is a **Gamma** distribution ($Gamma(\alpha,\beta)$) where the mean of $\gamma$ is $\frac{\alpha}{\beta}$, the variance is $\frac{\alpha}{\beta^2}$ and the modal value is $\frac{\alpha-1}{\beta}$.  The updated posterior distribution is $\text{Gamma}(\alpha+n\cdot \bar{y}, \beta+n)$ and the posterior predictive distribution is a negative binomial of the same form.  This is because for predictive purposes rather than exactly knowing $\gamma$ we have a range of possible $\gamma$ given by the updated distributions.  This mixture is now a negative binomial distribution

###### Example Counts-Based Updating

Lets say there is a number of adverse events (in a 30 day period) from a dietary supplement ($Y$) defined by $Pr(Y|\gamma)\sim\text{Poisson}(\gamma)$ where $\gamma$ is the average number of events over a fixed 30 day period.  Historically there are 2 events per month, based on about 10 months of experiences.  Based on this we should set the prior probability to be $Pr(Y)\sim\text{Gamma}(20,10)$ because the mean should be $\frac{\alpha}{\beta}=2=\frac{\alpha}{10}$

```{r poissoin-example}
historical.events <- 2
experiences <- 10
alpha <- historical.events*experiences
beta <- experiences

new.months <- 3
total.new.events <- 10
```

Now over the next `r new.months` months we observe `r total.new.events` total new events, a max likelihood estimate of `r total.new.events/new.months`.  This is now updated as $\text{Gamma}(\alpha+n\cdot \bar{y}, \beta+n)=\text{Gamma}(20+10,10+3)=\text{Gamma}(30,13)$.  This indicates that the mean number of events shifts from `r alpha/beta` [`r qgamma(c(0.025, 0.975), alpha, beta)`] to `r (alpha+total.new.events)/(beta+new.months)` [`r qgamma(c(0.025, 0.975), alpha+total.new.events, beta+new.months)`], the modal number of events from `r (alpha-1)/beta` to `r (alpha+total.new.events-1)/(beta+new.months)`.  This indicates the average number of events per thirty days. This means that the posterior probability of the true event rate or $\gamma>3$ is low (`r pgamma(3,alpha+total.new.events,beta+new.months,lower.tail=F)`). The variance moved from `r alpha/beta^2` to `r (alpha+total.new.events)/(beta+new.months)^2`.  Going forward, using a negative binomial distribution, the probability of observing over 3 events per 30 days is `r pnbinom(3, size = alpha+total.new.events, prob = (beta+new.months)/(beta+new.months+1), lower.tail = FALSE)`.

```{r poisson-visualization}
data.frame(observations = seq(0,10,by=0.1)) |>
  mutate(prior=dgamma(observations,alpha,beta),
         posterior =dgamma(observations,
                           alpha+total.new.events,
                           beta+new.months)) |>
  pivot_longer(c(prior,posterior),names_to="type",values_to="probability") |>
  ggplot(aes(y=probability,x=observations,col=type)) +
  geom_line() +
  theme_classic(base_size=16) +
  theme(legend.position=c(0.8,0.3))

data.frame(counts = seq(0,10,by=1)) |>
  mutate(prior= dnbinom(counts,
                            size = alpha,
                            mu = alpha/beta),
         posterior= dnbinom(counts,
                            size = alpha+total.new.events,
                            mu = (alpha+total.new.events) / (beta+new.months))) |>
  pivot_longer(c(prior,posterior),
               names_to="type",
               values_to="probability") |>
  ggplot(aes(y=probability,x=counts,fill=type)) +
  geom_bar(stat='identity',position='dodge') +
  theme_classic(base_size=16) +
  theme(legend.position=c(0.8,0.8)) +
  labs(title="Predictive Probability")
```

## Multiple Parameter Models

Most real world problems require us to estimate more than one parameter, unlike the examples above.  That being said we are generally only interested in one or two of them.

### Jeffries Priors

One special class of priors are Jeffreys priors.  These are defined by the likelihood function (normal, poisson, binomial, *et cetera*) than by our pre-existing intution or data.  This makes them more objective, but perhaps less Bayesian.  The Jeffrey's prior is denoted as $Pr(\theta) \propto \sqrt{I(\theta)}$ where $I(\theta)$ is the Fisher information of $\theta$.  This measures how sensitive the distribution is to changes in the parameter.  This is why it is based on the average of differential of the square of the log likelihood $(\frac{d}{d\theta}\cdot \log Pr(y|\theta))^2$.  Jeffreys priors are therefore uninformative with respect to the information geometry of the likelihood, rather than being flat over the parameter values themselves [@jeffreys1961theory].

#### Likelihoods and Jeffreys Priors for Common Models

| Likelihood | Parameter(s) | Likelihood $p(y \mid \theta)$ | Fisher Information $I(\theta)$ | Jeffreys Prior $\pi_J(\theta)$ | Notes |
|-----------|--------------|---------------------------------|---------------------------------|---------------------------------|-------|
| **Bernoulli** | $p \in (0,1)$ | $p^y (1-p)^{1-y}$ | $\frac{1}{p(1-p)}$ | $\pi(p) \propto \frac{1}{\sqrt{p(1-p)}}$ | Proper; Beta$(\frac{1}{2}, \frac{1}{2})$ |
| **Binomial** | $p \in (0,1)$ | $\binom{n}{y} p^y (1-p)^{n-y}$ | $\frac{n}{p(1-p)}$ | $\pi(p) \propto \frac{1}{\sqrt{p(1-p)}}$ | Same as Bernoulli (constant drops out) |
| **Poisson** | $\lambda > 0$ | $\frac{\lambda^y e^{-\lambda}}{y!}$ | $\frac{1}{\lambda}$ | $\pi(\lambda) \propto \lambda^{-1/2}$ | Improper; flat in $\sqrt{\lambda}$ |
| **Normal (mean only)** | $\mu \in \mathbb{R}$ | $\mathcal{N}(\mu, \sigma^2)$ | $\frac{1}{\sigma^2}$ | $\pi(\mu) \propto 1$ | Location parameter |
| **Normal (variance only)** | $\sigma > 0$ | $\mathcal{N}(\mu, \sigma^2)$ | $\frac{2}{\sigma^2}$ | $\pi(\sigma) \propto \frac{1}{\sigma}$ | Scale-invariant |
| **Normal (mean & variance)** | $(\mu, \sigma)$ | $\mathcal{N}(\mu, \sigma^2)$ | — | $\pi(\mu, \sigma) \propto \frac{1}{\sigma}$ | Joint Jeffreys prior |
| **Multinomial** | $\mathbf{p} = (p_1, \dots, p_k)$ | $\frac{n!}{\prod_i y_i!} \prod_i p_i^{y_i}$ | $\mathrm{diag}(1/p_i)$ | $\pi(\mathbf{p}) \propto \prod_i p_i^{-1/2}$ | Dirichlet$(\frac{1}{2}, \dots, \frac{1}{2})$ |

```{r jeffries-prior-visualizations}
p <- seq(0.001, 0.999, length.out = 500)

tibble(
  p = p,
  jeffreys = dbeta(p, 0.5, 0.5),
  uniform = dbeta(p, 1, 1),
  informative = dbeta(p, 14, 6)
) |>
  pivot_longer(-p, names_to = "prior", values_to = "density") |>
  ggplot(aes(x = p, y = density, color = prior)) +
  geom_line(linewidth = 1) +
  theme_classic(base_size = 6) +
  theme(legend.position=c(0.3,0.8)) +
  labs(
    title = "Priors for Binomial Probability p",
    x = expression(p),
    y = "Density"
  )-> binomial.plot


lambda_grid <- seq(0.01, 30, length.out = 500)

tibble(
  lambda = seq(0.01, 30, length.out = 500),
  jeffreys = lambda^(-1/2),          # unnormalized
  informative = dgamma(lambda, shape = 10, rate = 2)
) |>
  pivot_longer(-lambda, names_to = "type", values_to = "density") |>
  ggplot(aes(x = lambda, y = density, color = type)) +
  geom_line(linewidth = 1) +
  theme_classic(base_size = 6) +
  theme(legend.position=c(0.3,0.5)) +
  labs(
    title = "Jeffreys vs Informative Prior (Poisson)",
    x = expression(lambda),
    y = "Prior density \n(unnormalized for Jeffreys)"
  ) -> poisson.plot

mu <- seq(-5, 25, length.out = 500)

tibble(
  mu = mu,
  jeffreys = rep(1, length(mu)),                 # flat
  informative = dnorm(mu, mean = 10, sd = 2)
) |>
  pivot_longer(-mu, names_to = "prior", values_to = "density") |>
  ggplot(aes(x = mu, y = density, color = prior)) +
  geom_line(linewidth = 1) +
  theme_classic(base_size = 6) +
  theme(legend.position=c(0.8,0.5)) +
  labs(
    title = "Priors for Normal Mean (σ known)",
    x = expression(mu),
    y = "Density"
  ) -> norm.plot

library(cowplot)
plot_grid(plotlist=c(binomial.plot,poisson.plot,norm.plot),
          nrow=1)
```

#### Example Using Jeffrey's Priors 

Considering ultraprocessed foods, asking whether a food recommends the FDA recommendation for sodium in a single serving.  This is based on a survey of foods in a cafeteria.  The question (binomial) is whether a product exceeds or does not exceed the limit.

```{r jeffreys-setup}
products.sampled <- 12
exceeded <- 9

# based on 90% estimate
informed.alpha <- 90+1 #observed + 1
informed.beta <- 10+1 #not-observed +1
```

We sampled `r products.sampled` products, finding that `r exceeded` of them exceeded the limits.  This is denoted by $Binomial(n=`r products.sampled`,p)$.  For the binomial distribution, the Jeffreys prior for a binomial distribution is $Beta(\frac{1}{2},\frac{1}{2})$, which makes sense as we have no preferred parameterization.  The posterior probability in this case is $Beta(\frac{1}{2}+`r exceeded`,\frac{1}{2}+`r products.sampled`-`r exceeded`) = Beta(`r 0.5+ exceeded`,`r 0.5 + products.sampled-exceeded`)$.  The MLE is `r exceeded/products.sampled`.  We estimate the posterior mean (the percent of products that exceed limits) is `r (0.5+ exceeded)/(0.5+ exceeded+0.5 + products.sampled-exceeded)` $\pm$ [`r qbeta(c(0.025,0.975), shape1=0.5+ exceeded,shape2=0.5 + products.sampled-exceeded)`] with a modal value of `r (0.5+ exceeded-1)/(0.5+ exceeded+0.5 + products.sampled-exceeded-2)`. This is in comparason to a much more informed prior $Binomial(`r informed.alpha`,`r informed.beta`)$ which gives a posterior mean of `r (informed.alpha+ exceeded)/(informed.alpha+ exceeded+informed.beta + products.sampled-exceeded)` $\pm$ [`r qbeta(c(0.025,0.975), shape1=informed.alpha+ exceeded,shape2=informed.beta + products.sampled-exceeded)`].  These three distributions are shown in the figure showing that with a jeffreys prior vs a strong informed prior there is a substantial difference, with the jeffreys prior being broad and closer to the MLE and the strong prior only modifying the posterior estimate from that prior slightly.

```{r jeffreys-upf-example}
data.frame(p=seq(0.001, 0.999, length.out = 200)) |>
  mutate(jeffreys_prior=dbeta(p,0.5,0.5),
         jeffreys_posterior=dbeta(p,0.5+exceeded,0.5-exceeded+products.sampled),
         informed_prior=dbeta(p,informed.alpha,informed.beta),
         informed_posterior=dbeta(p,informed.alpha+exceeded,informed.beta-exceeded+products.sampled)) |> 
  pivot_longer(-p,names_sep="_",
               names_to=c("prior","type"),
               values_to='density') |>
  ggplot(aes(y=density,x=p,col=type,lty=prior)) +
  geom_line() +
    geom_vline(xintercept=exceeded/products.sampled, lty=2, col="grey") +
  annotate(
    geom  = "text",
    x     = exceeded/products.sampled-0.01,           # ← your chosen x-position
    y     = 12,            # ← your chosen y-position
    label = "Observed",
    hjust = 1,            # 0 = left justified (most important part)
    vjust = 0.5,          # optional: 0.5 = vertically centered
    size  = 5, color = "grey"
  ) +
  theme_classic(base_size=16) +
  theme(legend.position=c(0.2,0.7))
```

### Nuisance Parameters

Perhaps there are several parameters that are required in a model, but we only care about one, so think of a model with parameters $\theta=(\theta_1,\theta_2)$ of which we only care about $\theta_1$, that makes $\theta_2$ a nuisance parameter.  For example we may only care about the mean but not the variance. In modeling the posterior $Pr(\theta|y) \propto Pr(y|\theta)\cdot(Pr(\theta)$.  To get $\Pr(\theta_1|y)$, the marginal posterior, we integrate out $\theta_2$.  This can give you the marginal distribution of $\theta_1$.  Another approach is to use the Jeffreys priors and only keep terms that include $\theta_1$.  For example for a normal distribution, using a Jeffreys prior the posterior probability is $Pr(\mu,\sigma^2|y) \propto (\sigma^2)^{\frac{n}{2}-\frac{1}{2}}\cdot \exp\{-\frac{\sigma^2}{2}[SS+n(\mu-\bar{y})^2]\}$ where $n$ is the number of new observations $SS$ is the new sum of squares and $\bar{y}$ is the new mean.

### Using Jeffreys Prior to Define a Marginal Posterior

For a normal distribution the Jeffreys prior is $Pr(\mu,\sigma) \propto \frac{1}{\sigma}$ or $\tau$.  We can derive the posterior probability using that and the likelihood function for the normal distribution and multiplying by a Jeffreys prior.  Integrating out $\tau$ gives the marginal posterior for $\mu$ gives me the $t$ distribution with $n-1$ degrees of freedom or $\mu|y \sim t_{n-1}(\bar{y}, \frac{SS}{n\cdot(n-1)})$ with a marginal predictive posterior probability of $y_{new}|y \sim t_{n-1}(\bar{y},\frac{SS(n+1)}{n(n-1)})$

#### Example Using Jeffrey's Derived Marginal Probabilities

```{r jeffreys-marginal-setup}
observations <- c(4.2,5.1,6.3,5.4,4.8)
observations.ss <- sum((observations-mean(observations))^2)
library(extraDistr)
```

We observe the following values `r paste(observations,sep=",")`.  Based on this $n=`r length(observations)`$, $\bar{y}=`r mean(observations)`$ and $SS=`r sum((observations-mean(observations))^2)`$.  Based on this the probability that $Pr(5.0<\mu<5.5∣y)$ is based on the difference between the two t-distributions for 5 and 5.5 $t_{n-1}(\bar{y},\frac{SS}{n\cdot(n-1)})$ or `r plst(5.5,
  df    = length(observations) - 1,
  mu    = mean(observations),
  sigma = sqrt(observations.ss / (length(observations) * (length(observations) - 1)))
) -
plst(
  5.0,
  df    = length(observations) - 1,
  mu    = mean(observations),
  sigma = sqrt(observations.ss / (length(observations) * (length(observations) - 1)))
)` as visualized below:

```{r jeffreys-marginal-visualization}
n <- length(observations)
mu_hat <- mean(observations)
sigma_mu <- sqrt(observations.ss / (n * (n - 1)))

df_plot <- data.frame(
  values = seq(0, 10, by = 0.01)
) |>
  mutate(
    density = dlst(
      values,
      df    = n - 1,
      mu    = mu_hat,
      sigma = sigma_mu
    )
  )

ggplot(df_plot, aes(x = values, y = density)) +
  geom_line(linewidth = 1) +
  geom_area(
    data = subset(df_plot, values >= 5 & values <= 5.5),
    aes(x = values, y = density),
    fill = "grey70",
    alpha = 0.6
  ) +
  geom_vline(xintercept = c(5, 5.5), linetype = 2, color = "grey40") +
  theme_classic(base_size = 16) +
  labs(
    x = expression(mu),
    y = "Posterior density",
    title = expression(paste("Posterior for ", mu, " under Jeffreys prior"))
  )
```

The predictive probability of a new value being between 4.5 and 6 is given by the differnce between the distributions $t_{n-1}(\bar{y},\frac{SS(n+1)}{n(n-1)})$ or `r plst(6,
  df    = length(observations) - 1,
  mu    = mean(observations),
  sigma = sqrt(observations.ss * (length(observations) + 1) / (length(observations) * (length(observations) - 1)))
) -plst(4.5,
  df    = length(observations) - 1,
  mu    = mean(observations),
  sigma = sqrt(observations.ss * (length(observations) + 1) / (length(observations) * (length(observations) - 1)))
)`.  This is a wider range because the scale parameter is larger due to the uncertainty in predicting a new value `r sqrt(observations.ss / (length(observations) * (length(observations) - 1)))` vs `r sqrt(observations.ss * (length(observations) + 1) / (length(observations) * (length(observations) - 1)))`.  This is shown here:

```{r jeffreys-marginal-predictive-visualization}

sigma_mu <- sqrt(observations.ss / (length(observations) * (length(observations) - 1)))
sigma_mu.pred <- sqrt(observations.ss * (length(observations) + 1) / (length(observations) * (length(observations) - 1)))

data.frame(
  values = seq(0, 10, by = 0.01)
) |>
  mutate(
    posterior = dlst(
      values,
      df    = n - 1,
      mu    = mu_hat,
      sigma = sigma_mu
    ),
    predictive= dlst(
      values,
      df    = n - 1,
      mu    = mu_hat,
      sigma = sigma_mu.pred
    ),
  ) |>
  pivot_longer(c(posterior, predictive),
               names_to="type",
               values_to="density") |>
  ggplot(aes(x = values, y = density, color=type)) +
  geom_line()+
  theme_classic(base_size = 16) +
  theme(legend.position=c(0.8,0.8)) 
```

## Summary Tables

### Predictive Probability Distributions

These are always larger due to the uncertainty in new predictions than the posterior probabilities.  They are summarized here:

| Distribution | Conjugate Prior | Predictive Posterior |
| -- | -- | -- |
| Binomial | Beta | Beta-Binomial |
| Bernoulli | Beta | Beta-Bernoulli |
| Normal (known $\sigma$) | Normal |Normal |
| Normal (unknown $\sigma$) | Normal-Inverse Gamma |$t$ |
| Poisson | Gamma | Negative Binomial | 
| Multinomial | Dirichlet | Dirichlet-Multinomial |
 
### Summary of Conjugate Priors

| Distribution | Prior | Updated Posterior | Mean | Mode | Variance |
| --- | --- | --- | --- | --- | ----| 
| $Binomial(n,\theta)$ | $Beta(\alpha, \beta)$ | $Beta(\alpha + y_o, \beta+n-y_o)$ | $\frac{\alpha}{\alpha + \beta}$  | $\frac{\alpha-1}{\alpha + \beta - 2}$ |$\frac{\alpha \cdot \beta}{(\alpha + \beta)^2(\alpha + \beta + 1)}$ |
| $N(\mu,\sigma^2)$ known mean | $InvGamma(\alpha,\beta)$ | $InvGamma(\alpha + \frac{n}{2},\beta + \frac{SS}{2})$ | $\frac{\beta}{\alpha-1}$  | $\frac{\beta}{\alpha+1}$ |$\frac{\beta^2}{(\alpha-1)^2(\alpha-2)}$  |
|$N(\mu,\sigma^2)$ known variance | $N(\mu_o,\tau^2_o)$| $N\left(
\frac{\frac{\mu_0}{\tau_0^2} + \frac{n \cdot \bar{y}}{\sigma^2}}
{\frac{1}{\tau_0^2} + \frac{n}{\sigma^2}},
\;
\frac{1}{\frac{1}{\tau_0^2} + \frac{n}{\sigma^2}}
\right)$ | $\mu$ | $\mu$ | $\sigma^2$|
| $N(\mu,\sigma^2)$ | $N-InvGamma(\mu_0, \kappa_0,\alpha_0,\beta_0)$ | $\mu|y ~\sim t(\frac{\kappa_0 \cdot \mu_0 + n \cdot \bar{y}}{\kappa_0+n},\sqrt{\frac{\beta_0+\frac{SS}{2}+\frac{\kappa_0 \cdot n}{2\kappa_0+n}\cdot{(\bar{y}-\mu_0})^2}{\alpha_0+\frac{n}{2} \cdot \kappa_0+n}})$$\sigma^2∣y∼InvGamma(\alpha_0+\frac{n}{2},\beta_0+\frac{SS}{2}+\frac{\kappa_0 \cdot n}{2\kappa_0+n}\cdot{(\bar{y}-\mu_0})^2)$ | $\mu$ \ $\frac{\beta}{\alpha-1}$| $\mu$ \ $\frac{\beta}{\alpha+1}$ | $\frac{\beta}{(\alpha-1)\cdot\kappa}$ $\frac{\beta^2}{(\alpha-1)^2\cdot(\alpha-2)}$|
| $\text{Poisson}(\gamma)$ | $\text{Gamma}(\alpha,\beta)$ | $\text{Gamma}(\alpha+n\cdot \bar{y}, \beta+n)$ | $\frac{\alpha}{\beta}$ | $\frac{\alpha-1}{\beta}$ | $\frac{\alpha}{\beta^2}$ | 
| $\text{Multinomial}(N,\pi)$ | $\text{Dirichlet}(\alpha+y_1...\alpha_k+y_k)$| $\frac{\alpha_i+y_i}{\sum_j(\alpha_j+y_j)}$ | $\frac{\alpha_i}{\alpha_o}$| $\frac{\alpha_i-1}{\alpha_o-K}$| $\frac{\bar{\alpha_i}(1-\bar{\alpha_i})}{\alpha_o+1}$ |

In these tables:

- $\alpha$ and $\beta$ vary depending on the model:
    - For inverse gamma distributions $\alpha=\text{successes}+1$ and $\beta=\text{failures}+1$. 
    - For normal distributions if using existing data, one heuristic could be $\alpha=\frac{\kappa}{2}$ (prior degrees of freedom) and $\beta=\frac{\sigma^2\cdot\kappa}{2}$ (prior total sum of squares)
    - For poisson distributions, $\beta$ is the number of prior observations and $\alpha$ is the sum of events from those prior observations.
    - For dirichlet, $alpha_i$ is the prior pseudo-count for category $i$ and $alpha_o$ is $\alpha_0 = \sum\limits_{i=1}^{K}$, the sum of all the alphas from 1 to k, the prior sample size.
- $n$ is the number of new trials.
- $y_o$ is the number of successes.
- $SS$ is the sum of squares of the new data.
- $\mu$ is the (known in some cases) mean, or the prior pseudo observation.
- $\sigma^2$ is the (known in some cases) variance.
- $\bar{y}$ is the mean of the new data.
- $\kappa$ is the strength of the prior belief in $\mu$, or the effective sample size

### Posterior and Predictive Probabilities with Jeffreys Priors

| Distribution                | Likelihood                                              | Jeffreys prior                                          | Posterior (parameters)                                        | Predictive (new data)                                                     |
| --------------------------- | ------------------------------------------------------- | ------------------------------------------------------- | ------------------------------------------------------------- | ------------------------------------------------------------------------- |
| Bernoulli / Binomial        | $y \sim \mathrm{Binomial}(n,p)$                         | $p(p)\propto [p(1-p)]^{-1/2}$                           | $p\mid y \sim \mathrm{Beta}(y+1/2,;n-y+1/2)$                  | $y_{\text{new}}\mid y \sim \mathrm{Beta\text{-}Binomial}$                 |
| Poisson                     | $y_i \sim \mathrm{Poisson}(\lambda)$                    | $p(\lambda)\propto \lambda^{-1/2}$                      | $\lambda\mid y \sim \mathrm{Gamma}(\sum y_i+1/2,;n)$          | $y_{\text{new}}\mid y \sim \mathrm{Neg.\ Binomial}$                       |
| Multinomial                 | $\mathbf y \sim \mathrm{Multinomial}(n,\boldsymbol\pi)$ | $\boldsymbol\pi \sim \mathrm{Dirichlet}(1/2,\dots,1/2)$ | $\boldsymbol\pi\mid y \sim \mathrm{Dirichlet}(y_1+1/2,\dots)$ | $\mathbf y_{\text{new}}\mid y \sim \mathrm{Dirichlet\text{-}Multinomial}$ |
| Normal (μ unknown, σ known) | $y_i\sim\mathcal N(\mu,\sigma^2)$                       | $p(\mu)\propto 1$                                       | $\mu\mid y \sim \mathcal N(\bar y,\sigma^2/n)$                | $y_{\text{new}}\mid y \sim \mathcal N(\bar y,\sigma^2(1+1/n))$            |
| Normal (μ known, σ unknown) | $y_i\sim\mathcal N(\mu,\sigma^2)$                       | $p(\sigma^2)\propto 1/\sigma^2$                         | $\sigma^2\mid y \sim \mathrm{Inv\text{-}Gamma}(n/2,;SS/2)$    | $y_{\text{new}}\mid y \sim t_n(\mu,\sqrt{SS/n})$                          |
| Normal (μ, σ unknown)       | $y_i\sim\mathcal N(\mu,\sigma^2)$                       | $p(\mu,\sigma^2)\propto 1/\sigma^2$                     | $\mu\mid y \sim t_{n-1}(\bar y,;s/\sqrt n)$                   | $y_{\text{new}}\mid y \sim t_{n-1}(\bar y,;s\sqrt{1+1/n})$                |


## References

::: {#refs}
:::

## Session Information

```{r session-information}
sessionInfo()
```
