---
title: "Correlations and Linear Models"
author: "Dave Bridges"
date: '2025-04-29'
editor: source
format:
  html:
    toc: true
    toc-location: right
    keep-md: true
theme: journal
execute:
  echo: true
  warning: false
  message: false
bibliography: references.bib
---

```{r libraries}
library(tidyverse)
library(moderndive)
library(broom)
library(knitr)
```

We will use the `house_prices` dataset from the `moderndive` package, which contains sale prices and characteristics of houses in King County, Washington. Here we explore how the square feet of living space (`sqft_living`) relates to sale `price`.

```{r color-scheme}
color_scheme <- c("#00274c", "#ffcb05")
```

## Visualizing a Relationship

The first step is always to plot the two variables against each other. Here we overlay both a smoothed (loess) curve and a linear fit to see whether a straight line is a reasonable description of the relationship.

```{r price-sqft-living}
ggplot(house_prices, aes(y = price, x = sqft_living)) +
  geom_point(alpha = 0.1) +
  geom_smooth(method = "loess", color = color_scheme[2]) +
  geom_smooth(method = "lm",    color = color_scheme[1]) +
  theme_classic(base_size = 16)
```

The relationship appears approximately linear (blue line), though the loess curve (maize) suggests a slight upward bend at larger values, indicating some non-linearity. We will address this in the [Non-Linear Relationships](#non_linear) section.

## Quantifying the Relationship: Correlation

Before fitting a regression model it is useful to quantify the strength of the association with a correlation coefficient. There are two main choices.

### Pearson Correlation

Pearson's *r* measures the strength of the **linear** association. Inference (the t-test for H₀: ρ = 0) relies on bivariate normality — both variables should be approximately normally distributed. With large samples the Central Limit Theorem makes the test robust to moderate departures, but Pearson *r* is sensitive to outliers and is attenuated by non-linearity.

```{r pearson}
cor.test(house_prices$price, house_prices$sqft_living, method = "pearson") |>
  tidy() |>
  kable(digits = 3, caption = "Pearson correlation between price and living area")
```

### Spearman Correlation

Spearman's ρ is a rank-based measure of **monotonic** association. It does not assume normality and is robust to outliers and non-linearity. It is the better default when the data are skewed or when the relationship may be monotonic but not strictly linear — both of which apply here.

```{r spearman}
cor.test(house_prices$price, house_prices$sqft_living, method = "spearman") |>
  tidy() |>
  kable(digits = 3, caption = "Spearman correlation between price and living area")
```

Spearman ρ will generally differ from Pearson *r* when the data are skewed or the relationship is non-linear. A noticeably higher Spearman ρ suggests the relationship is monotonic but not strictly linear — which matches what the loess curve shows.

## Describing the Linear Relationship with Regression

A linear model provides both the **slope** (how much price changes per additional square foot) and **R²** (the proportion of variance in price explained by living area). We extract these with `tidy()` (coefficients) and `glance()` (overall fit), adding confidence intervals throughout.

```{r linear-model}
lm.1 <- lm(price ~ sqft_living, data = house_prices)

tidy(lm.1, conf.int = TRUE) |>
  kable(digits = 2, caption = "Coefficient estimates with 95% confidence intervals")

glance(lm.1) |>
  kable(digits = 3, caption = "Overall model fit")
```

We estimate that price increases by \$`r round(lm.1$coefficients['sqft_living'], 0)` (95% CI: \$`r round(confint(lm.1)['sqft_living', 1], 0)` – \$`r round(confint(lm.1)['sqft_living', 2], 0)`) per square foot of living space. The model explains `r round(summary(lm.1)$r.squared * 100, 1)`% of the variance in price (R² = `r round(summary(lm.1)$r.squared, 3)`). For simple regression with one predictor, R² equals the squared Pearson correlation. When comparing models with different numbers of predictors, use adjusted R² from `glance()` instead, which penalises additional parameters.

### Confidence Intervals vs. Prediction Intervals

It is important to distinguish two types of interval around the fitted line:

- A **confidence interval** describes uncertainty in the *mean* response at a given x — how well we know where the regression line is.
- A **prediction interval** describes uncertainty for a *new individual observation* at a given x — it is always wider because it must also account for the residual variance around the line.

For applied questions (e.g. "what will this specific house sell for?"), prediction intervals are what matter.

```{r intervals}
new_sizes <- data.frame(sqft_living = c(1000, 2000, 3000))

bind_cols(
  new_sizes,
  predict(lm.1, newdata = new_sizes, interval = "confidence") |>
    as_tibble() |> rename(ci_lwr = lwr, ci_upr = upr),
  predict(lm.1, newdata = new_sizes, interval = "prediction") |>
    as_tibble() |> select(pi_lwr = lwr, pi_upr = upr)
) |>
  kable(digits = 0,
        caption = "Fitted values with 95% confidence intervals (CI) and prediction intervals (PI)")
```

The prediction intervals are substantially wider than the confidence intervals, illustrating that even a well-fitting model has considerable uncertainty for individual predictions.

### Testing the Assumptions of the Model

For a linear model there are four assumptions:

- **Linearity**: The relationship between predictor and outcome is linear — one unit change in x produces a constant change in y.
- **Independence**: Residuals are independent of each other and of the predictor.
- **Homoscedasticity**: Residuals have constant variance across all fitted values.
- **Normality**: Residuals are approximately normally distributed.

```{r lm-assumptions}
par(mfrow = c(2, 2))
plot(lm.1)
```

#### Interpreting These Diagnostic Plots

- **Residuals vs. Fitted**: Residuals should scatter randomly around zero with no pattern. A cone shape (widening spread) indicates heteroscedasticity; a curve indicates non-linearity.
- **Normal Q-Q**: Residuals should fall on the diagonal reference line. Deviations at the upper tail (points curve above the line) indicate a heavy right tail or right skewness. An S-shape indicates the tails are lighter than normal (platykurtic); a reverse S-shape indicates heavier tails (leptokurtic).
- **Scale-Location**: The square root of standardised residuals should be approximately flat. An upward slope indicates heteroscedasticity.
- **Residuals vs. Leverage**: Points outside Cook's distance dashed lines are influential — their removal would substantially change the fitted coefficients.

#### Assessing Residual Normality

A Q-Q plot of model residuals is the primary visual diagnostic for normality. Points should fall approximately on the diagonal; systematic departures indicate the residuals are not normally distributed.

```{r qq-residuals}
tibble(residuals = residuals(lm.1)) |>
  ggplot(aes(sample = residuals)) +
  stat_qq(alpha = 0.1) +
  stat_qq_line(color = color_scheme[1]) +
  labs(
    title = "Normal Q-Q plot of model residuals",
    x     = "Theoretical normal quantiles",
    y     = "Observed residual quantiles"
  ) +
  theme_classic(base_size = 14)
```

A Shapiro-Wilk test can supplement the visual check, but note that with large datasets it has very high power — it will reject normality for trivial departures that have no practical effect on inference. Treat the Q-Q plot as the primary diagnostic and the p-value as a supplementary signal.

```{r shapiro-test}
# Shapiro-Wilk requires n ≤ 5000; sample from the full dataset
set.seed(42)
bind_rows(
  sample(house_prices$price,       5000) |> shapiro.test() |> tidy() |> mutate(variable = "price"),
  sample(house_prices$sqft_living, 5000) |> shapiro.test() |> tidy() |> mutate(variable = "sqft_living"),
  sample(residuals(lm.1),          5000) |> shapiro.test() |> tidy() |> mutate(variable = "residuals")
) |>
  relocate(variable) |>
  kable(digits = 3, caption = "Shapiro-Wilk tests (n = 5000 sample; see text on large-n interpretation)")
```

Here both variables and the residuals are significantly non-normal, and the diagnostic plots confirm heteroscedasticity (cone-shaped residuals vs. fitted, upward-sloping scale-location plot) and a heavy upper tail (Q-Q points lift above the line at high quantiles). This combination — right-skewed outcome, heteroscedasticity, slight non-linearity — is typical of price data and suggests a log transformation of the response is worth trying (see below).

#### What to Do if Assumptions Are Not Met

- **Non-linearity**: Apply a transformation (log, square root) to the predictor or response, add polynomial or spline terms, or use a generalised additive model (GAM). See [Non-Linear Relationships](#non_linear).
- **Heteroscedasticity**: Log-transform the response (often fixes both heteroscedasticity and non-linearity for skewed outcomes); use weighted least squares; or use heteroscedasticity-consistent (sandwich) standard errors via `lmtest::coeftest()` with `sandwich::vcovHC()`.
- **Non-normality of residuals**: Transformation is usually the first step. With large n, the CLT means non-normal residuals have little effect on coefficient inference, but they do affect prediction intervals. Robust regression (`MASS::rlm()`) or quantile regression are alternatives.
- **Non-independence**: For clustered or repeated-measures data, use mixed-effects models (`lme4::lmer()`). For time series, use ARIMA or include lagged terms.

## Evaluating and Modifying Models for Non-Linear Relationships {#non_linear}

Given the diagnostic evidence of non-linearity and heteroscedasticity, we can try a log transformation of the response and compare it against the original model and a polynomial extension.

```{r log-model}
lm.log <- lm(log(price) ~ sqft_living, data = house_prices)

tidy(lm.log, conf.int = TRUE) |>
  kable(digits = 4, caption = "Log-price model coefficients")

par(mfrow = c(2, 2))
plot(lm.log)
```

The slope on the log scale is a proportional effect. The exact percentage change in price per unit increase in `sqft_living` is $(\exp(\beta) - 1) \times 100$. The approximation $\beta \times 100$ is adequate only when $|\beta| \lesssim 0.05$; for larger coefficients the approximation understates the true effect.

```{r log-slope-interpretation}
beta_log <- coef(lm.log)["sqft_living"]
tibble(
  slope_log_scale      = beta_log,
  pct_change_exact     = (exp(beta_log) - 1) * 100,
  pct_change_approx    = beta_log * 100
) |>
  kable(digits = 4,
        caption = "Exact vs approximate percentage-change interpretation of log-scale slope")
```

The diagnostic plots should now show more uniform spread and a straighter Q-Q line if the transformation was appropriate.

We can also fit a polynomial model on the original scale (valid AIC comparison since the response is the same). We use `poly(sqft_living, 2)` rather than `sqft_living + I(sqft_living^2)` because `poly()` generates orthogonal polynomial terms, eliminating the collinearity between x and x² that inflates standard errors with raw polynomial coding.

```{r poly-model}
lm.poly <- lm(price ~ poly(sqft_living, 2), data = house_prices)

tidy(lm.poly, conf.int = TRUE) |>
  kable(digits = 2, caption = "Orthogonal polynomial model coefficients")
```

### Comparing Models with AIC and BIC

AIC and BIC penalise model complexity and can identify which model fits the data best relative to its degrees of freedom. **Models must have the same response variable to be compared** — so we compare the log-response models with each other, and the original-scale models with each other, but not across the two response scales.

```{r model-comparison}
# Original-scale models: linear vs. quadratic
AIC(lm.1, lm.poly) |>
  rownames_to_column("model") |>
  kable(digits = 1, caption = "AIC comparison: linear vs. quadratic (original scale)")

BIC(lm.1, lm.poly) |>
  rownames_to_column("model") |>
  kable(digits = 1, caption = "BIC comparison: linear vs. quadratic (original scale)")
```

Lower AIC/BIC indicates better fit adjusted for model complexity. A difference of >2 is considered meaningful; >10 is strong evidence in favour of the lower-scoring model.

For **nested models** — where one model is a special case of another — a likelihood ratio test via `anova()` is more powerful than AIC and directly tests whether the extra terms improve fit significantly. AIC and BIC are more appropriate for comparing non-nested models.

```{r nested-lrt}
anova(lm.1, lm.poly) |>
  tidy() |>
  kable(digits = 3, caption = "Likelihood ratio test: linear vs. quadratic (nested comparison)")
```

Alternatively, spline terms offer a flexible non-linear fit without specifying the polynomial degree:

```{r spline-model}
library(splines)
lm.spline <- lm(price ~ ns(sqft_living, df = 4), data = house_prices)

AIC(lm.1, lm.poly, lm.spline) |>
  rownames_to_column("model") |>
  kable(digits = 1, caption = "AIC comparison across three model structures")
```

The natural spline with 4 degrees of freedom allows up to 3 bends in the curve. Re-examine the residual plots after fitting to confirm improvement in linearity.

## Bayesian Approach

For an alternative to null hypothesis significance testing, a Bayesian approach estimates the posterior distribution of the slope and R² directly. See [the Bayesian tutorial](https://bridgeslab.github.io/Lab-Documents/Experimental%20Policies/bayesian-analyses.html) for background.

We set an illustrative informative prior based on published estimates (here ~\$1,663/sq ft with SD \$602) and use a Student-*t* likelihood, which is more robust to the outliers and heteroscedasticity diagnosed above.

```{r brms-linear-regression}
#| output: false
library(brms)

# directory for cached model fits — brms reuses these on re-render
dir.create("fits", showWarnings = FALSE)

housing.priors <- prior(normal(1663, 602), class = b, coef = sqft_living)
brm.1 <- brm(
  price ~ sqft_living,
  data         = house_prices,
  prior        = housing.priors,
  family       = student(),
  sample_prior = TRUE,
  file         = "fits/housing-prices",
  file_refit   = "on_change"
)
```

```{r linear-analysis}
prior_summary(brm.1) |>
  kable(caption = "Prior summary")

library(broom.mixed)
tidy(brm.1) |>
  kable(digits = 2, caption = "Posterior estimates for price vs. square feet of living space")

plot(brm.1)

hypothesis(brm.1, "sqft_living > 0")   # is the slope positive?
hypothesis(brm.1, "sqft_living = 1663") # does the slope match the prior mean?
```

```{r posterior-slope}
as_draws_df(brm.1) |>
  ggplot(aes(x = b_sqft_living)) +
  geom_density(fill = color_scheme[2]) +
  labs(
    x     = "Slope: price per square foot ($)",
    y     = "",
    title = "Posterior distribution of slope"
  ) +
  theme_classic(base_size = 16)
```

The Bayesian estimate (\$`r round(fixef(brm.1)['sqft_living', 'Estimate'], 0)` ± `r round(fixef(brm.1)['sqft_living', 'Est.Error'], 0)`) is similar to the OLS estimate (\$`r round(lm.1$coefficients['sqft_living'], 0)` ± `r round(summary(lm.1)$coefficients['sqft_living', 'Std. Error'], 0)`), though the Student-*t* likelihood downweights influential high-value outliers slightly.

```{r r2-calculations}
bayes_R2(brm.1) |>
  kable(digits = 3, caption = "Bayesian R²")

bayes_R2(brm.1, summary = FALSE) |>
  as_tibble() |>
  ggplot(aes(x = R2)) +
  geom_density(fill = color_scheme[2]) +
  labs(x = "R²", y = "", title = "Posterior distribution of R²") +
  theme_classic(base_size = 16)
```

A posterior predictive check confirms whether the model captures the data-generating process. The lingering non-linearity means that even the Bayesian model with a robust likelihood will show some systematic mis-fit at high fitted values.

```{r brms-pp-check}
pp_check(brm.1, type = "scatter_avg")
pp_check(brm.1, type = "dens_overlay")
```

## References

::: {#refs}
:::

## Session Information

```{r session-information}
#| echo: false
sessionInfo()
```
